---
title: "PCWD values from ModE-Sim vs ERA5Land output"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(ggplot2)
library(cwd)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(maps)
library(raster)
library(rnaturalearth)
library(sf)
library(rgdal)
library(sp)
library(RColorBrewer)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
library(geoprocessoR)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
library(DescTools) #for GINI calculation
library(ggsci)
```
### Check PET and PRECIP data are similar

```{r}
#read in one ensemble member data from ModE-Sim
ModE_Sim_rds <- readRDS("/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m002_tidy/02_pcwd_1850/ModESim_pcwd_LON_-071.250.rds")
lon_ModESim <- ModE_Sim_rds$lon[1]

#read in one year of ERA5 regridded data: 
input_ERA5_raw <- "/storage/research/giub_geco/data_2/scratch/phelpap/ERA5Land_1950-2024/ERA5Land_regridded/ERA5Land_UTCDaily.tp_pev.1968.nc"
nc_pev <- nc_open(input_ERA5_raw)
pev = ncvar_get(nc_pev, varid="tot_pev") #in units of m
precip = ncvar_get(nc_pev, varid= "tot_tp") # in units of m
lon = ncvar_get(nc_pev, varid="lon")
lat = ncvar_get(nc_pev, varid="lat") # --- will contain less latitudes as only contains Land data
time_ERA5 = ncvar_get(nc_pev, varid="valid_time")
nc_close(nc_pev)

#from ERA5 data only select the same longitude : ERA5 now has year 1968 and only one longitude
lon_indices <- which(lon == lon_ModESim)
pev_ERA5LONfiltered <- pev[lon_indices,,]  ## has NAs where there is no land
# Remove rows with NA values
pev_ERA5LONfiltered_no_na <- na.omit(pev_ERA5LONfiltered)
# Check the structure of the cleaned dataset
str(pev_ERA5LONfiltered_no_na)

precip_ERA5LONfiltered <- precip[lon_indices,,] ## has NAs where there is no land
# Remove rows with NA values
precip_ERA5LONfiltered_no_na <- na.omit(precip_ERA5LONfiltered)
# Check the structure of the cleaned dataset
str(precip_ERA5LONfiltered_no_na)

```


```{r}
### restructure ModE-Sim data of rds format to match the ERA5 data
#from ModESim data only retain year 1968 for pet and precip 
library(dplyr)
library(purrr)

# Function to filter each nested df for 1968 and retain only 'pet' and 'precipitation'
filter_1968_data <- function(data_list) {
  data_list$df <- data_list$df %>%
    filter(lubridate::year(date) == 1968) %>%  # Keep only rows from 1968
    dplyr::select(date, pet, precipitation)  # Retain only relevant columns
  
  data_list$inst <- NULL  # Remove the 'inst' list
  
  return(data_list)  # Return modified list
}

# Apply the function to all elements in the tibble
filtered_tibble <- ModE_Sim_rds %>%
  mutate(data = purrr::map(data, filter_1968_data))  # Apply to each list in 'data'

# Check for NAs in the pev_ERA5LONfiltered dataset (assuming it's a matrix)
# Convert matrix to data.frame or tibble for easier manipulation
pev_ERA5LONfiltered_df <- as.data.frame(pev_ERA5LONfiltered)

# Find the latitudes (rows) with NAs
na_latitudes <- which(rowSums(is.na(pev_ERA5LONfiltered_df)) > 0)

# Now you have the indices of the latitudes that contain NAs
# To filter the filtered_tibble based on latitudes without NAs, first extract latitudes from your filtered_tibble
# Assuming your filtered_tibble contains a "lat" column (or similar), we can do:

# Filter out rows based on the indices in na_latitudes
filtered_tibble_no_na <- filtered_tibble[-na_latitudes, ]

# Check the structure to ensure latitudes are correctly removed
str(filtered_tibble_no_na)
# Display the result to verify the filtering
filtered_tibble_no_na
```
Both the ERA5 pet and precip data (precip_ERA5LONfiltered_no_na and pev_ERA5LONfiltere_no_na) and the ModESim data (filtered_tibble_no_na) contain only data for one longitude, latitudes where there are no nas and only for the year 1968. We can now check the distribution of both
```{r}
#melt all data into vectors:

#### MOD E Sim data
# Extract all precipitation data
precipitation_data <- filtered_tibble_no_na %>%
  mutate(precipitation = purrr::map(data, ~ .x$df$precipitation)) %>%
  pull(precipitation)

# If you want the data as a single vector (combining all the precipitation values)
all_precipitation_ModESim <- unlist(precipitation_data)

# Extract all pet data
pet_data <- filtered_tibble_no_na %>%
  mutate(pet = purrr::map(data, ~ .x$df$pet)) %>%
  pull(pet)

# If you want the data as a single vector (combining all the pet values)
all_pet_ModESim <- unlist(pet_data)

### ERA 5 Land data
#convert into mm
precip_ERA5LONfiltered_conv <- precip_ERA5LONfiltered_no_na*1000 #from m to mm
pev_ERA5LONfiltered_conv <- pev_ERA5LONfiltered_no_na*1000*-1

all_precipitation_ERA5Land <- as.vector(precip_ERA5LONfiltered_conv)
all_pet_ERA5Land <- as.vector(pev_ERA5LONfiltered_conv)


##### compare data
summary(all_pet_ModESim)
summary(all_pet_ERA5Land)
summary(all_precipitation_ModESim)
summary(all_precipitation_ERA5Land)


```

plot density curves of the values
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Assuming you have the second dataset, for example:
# all_precipitation_ERA5Land <- ... (your data)
# Let's also assume the second dataset is stored in `all_precipitation_other`.

# Combine both datasets into a single data frame for plotting
combined_data <- tibble(
  precipitation = c(all_precipitation_ERA5Land, all_precipitation_ModESim),
  dataset = rep(c("ERA5Land", "ModESim"), c(length(all_precipitation_ERA5Land), length(all_precipitation_ModESim)))
)

# Plot the density curves
ggplot(combined_data, aes(x = precipitation, fill = dataset)) +
  geom_density(alpha = 0.4) +  # alpha controls transparency
  scale_fill_manual(values = c("blue", "red")) +  # You can customize colors here
  labs(
    title = "Density Curves of Precipitation Data (1968, one longitude)",
    x = "Precipitation [mm/d]",
    y = "Density",
    fill = "Dataset"
  ) +
  theme_minimal() +  # Use minimal theme
  theme(
    legend.position = "top"  # Position the legend at the top
  )

# Combine both datasets into a single data frame for plotting
combined_data <- tibble(
  pet = c(all_pet_ERA5Land, all_pet_ModESim),
  dataset = rep(c("ERA5Land", "ModESim"), c(length(all_pet_ERA5Land), length(all_pet_ModESim)))
)

# Plot the density curves
ggplot(combined_data, aes(x = pet, fill = dataset)) +
  geom_density(alpha = 0.4) +  # alpha controls transparency
  scale_fill_manual(values = c("blue", "red")) +  # You can customize colors here
  labs(
    title = "Density Curves of PET Data (1968, one longitude)",
    x = "pet [mm/d]",
    y = "Density",
    fill = "Dataset"
  ) +
  theme_minimal() +  # Use minimal theme
  theme(
    legend.position = "top"  # Position the legend at the top
  )

```




### ERA5Land data

Read in and select only the overlapping years (1968-2009)
```{r}
#lat lon and time info from netcdf files
##read in data
input_file_ERA5 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ERA5Land_1950-2024/04_result/PCWD_ANNMAX.nc"

nc_pwcd_ERA5 <- nc_open(input_file_ERA5)
pcwd_annmax_ERA5 = ncvar_get(nc_pwcd_ERA5, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_ERA5, varid="lon")
lat = ncvar_get(nc_pwcd_ERA5, varid="lat") # --- will contain less latitudes as only contains Land data
time_ERA5 = ncvar_get(nc_pwcd_ERA5, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates_ERA5 <- reference_date + time_ERA5

# Filter data to retain only years up to 2009
years <- as.numeric(format(time_dates_ERA5, "%Y"))
keep_indices <- which(years <= 2009)

# Subset time and data arrays
time_dates_ERA5 <- time_dates_ERA5[keep_indices]
pcwd_annmax_ERA5 <- pcwd_annmax_ERA5[,,keep_indices]  # Assuming time is the 3rd dimension


nc_close(nc_pwcd_ERA5)

```



### ModE-Sim data 

Read in lat, lon and time info: 
```{r}
#lat lon and time info from netcdf files
##read in data
input_file_1850 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850/PCWD_ANNMAX.nc"

nc_pwcd_1850 <- nc_open(input_file_1850)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat") 
time_1850 = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates_1850 <- reference_date + time_1850

# Filter data to retain only years up to 2009
years <- as.numeric(format(time_dates_1850, "%Y"))
keep_indices <- which(years >= 1968)

# Subset time and data arrays
time_dates_1850 <- time_dates_1850[keep_indices]
pcwd_annmax_1850 <- pcwd_annmax_1850[,,keep_indices]  # Assuming time is the 3rd dimension

nc_close(nc_pwcd_1850)
```

aggregate files and select only the overlapping years (1968-2009): 
```{r}
#### 1850 set: 
# Define the base path - have to think of way to seperate sets once there is data!!
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "04_result_1850/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_1850 <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Convert time variable to actual dates (assuming "days since 2001-01-01")
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time_1850

# Find indices corresponding to years >= 1968
years <- as.numeric(format(time_dates, "%Y"))
keep_indices <- which(years >= 1968)

# Loop through all the target files and filter data
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Subset data to retain only years from 1968 onwards
  data_filtered <- data[,,keep_indices]
  
  # Store the filtered data in the list
  data_by_ensemble_1850[[member]] <- data_filtered
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_1850))  # Check stored ensemble members
print(dim(data_by_ensemble_1850[[1]]))  # Check dimensions of the data for the first ensemble member

```

Compare global distribution of data sets
```{r}
#Calculate epoch mean of ModESim: 
ensemble_array <- simplify2array(data_by_ensemble_1850)  # Shape: [192, 96, 42, 20]

# Compute the ensemble mean across the 3rd dimension (time)
ensemble_means_1850 <- apply(ensemble_array, c(1, 2, 4), mean, na.rm = TRUE)  # Shape: [192, 96, 20]
dim(ensemble_means_1850)

#Calculate epoch mean for ERA5:
ensemble_array <- simplify2array(pcwd_annmax_ERA5)  # Shape: [192, 96, 42]

# Compute the ensemble mean across the 3rd dimension (time)
means_ERA5 <- apply(ensemble_array, c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]
dim(means_ERA5)

```
```{r}
library(tidyverse)

# Reshape the ensemble data: it’s a 3D array with dimensions (lon, lat, ensemble)
ens_df <- as.data.frame.table(ensemble_means_1850, responseName = "value") %>%
  rename(lon = Var1, lat = Var2, ensemble = Var3) %>%
  mutate(ensemble = factor(ensemble))

# Flatten the ERA5 means matrix and create a data frame.
era5_df <- data.frame(value = as.vector(means_ERA5)) %>%
  mutate(dataset = "ERA5")

# Create the density plot
ggplot() +
  # Plot each ensemble member with a different colour
  geom_density(data = ens_df, aes(x = value, group = ensemble, colour = ensemble), alpha = 0.5) +
  # Overlay the ERA5 density in a distinct colour and line width
  #geom_density(data = era5_df, aes(x = value), colour = "black", linewidth = 1.5) +
  labs(title = "Density Plot Comparison",
       x = "Value",
       y = "Density",
       colour = "Ensemble Member") +
  theme_minimal()
```



```{r}
#plot ERA5Land data to validate
# Convert first year's data into a raster object
r <- raster(t(means_ERA5), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
#r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 0.1,0.15, 0.2, 0.3, 1, 2)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
custom_colors <- c("#004f5a",  # New color: a darker turquoise at the beginning
                   "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", "#e8b249")

# Define the numeric labels for the breaks
break_labels <- c(as.character(scales::label_number(accuracy = 1)(custom_bins[-length(custom_bins)])), "Ocean")

# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania


# Define the ggplot2 object with the gradient scale
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "mean PCWD (mm)", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "PCWD in 1850", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD (mm)", reverse = TRUE))  # Adjust legend for better presentation

```
Compute regional dataset for ERA5Land data: 

```{r}
library(dplyr)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
library(geoprocessoR)
library(terra)
library(ncdf4)
library(rnaturalearth)
library(sf)

### read in IPCC Region information:
#Load reference regions and coastlines:

load("/storage/homefs/ph23v078/Reference_regions/IPCC-WGI-reference-regions-v4_R.rda", verbose = TRUE)

#simplify this object by converting it to a SpatialPolygons class object (i.e., only the polygons are retained and their attributes discarded):
refregions <- as(IPCC_WGI_reference_regions_v4, "SpatialPolygons")

# List of regions to loop over --- excludes ocean basins
regions <- c("GIC", "NWN", "NEN", "WNA", "CNA", "ENA", "NCA", "SCA", "CAR", "NWS",
             "NSA", "NES", "SAM", "SWS", "SES", "SSA", "NEU", "WCE", "EEU", "MED",
             "SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG", "RAR", "WSB",
             "ESB", "RFE", "WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "NAU",
             "CAU", "EAU", "SAU", "NZ")
# regions <- c("GIC")

########## for ERA5Land to 1849, set 1 ###########################
#as before but loop over all regions also, saving everything in an array
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ERA5Land_1950-2024"
folders <- list.files(path, full.names = TRUE, pattern = "04_result") #annmax files

# Initialize a list to store results for all regions
regional_results_ERA5Land <- list()

# Loop over all regions
for (region in regions) {
  # Extract the spatial object corresponding to the current region
  region_object <- refregions[c(region)]

  # Check if the subset is valid
  if (is.null(region_object)) {
    warning(paste("Region not found:", region))
    next
  }

  # Initialize a list to store spatial averages for each file in the current region
  spatial_avg_list <- list()

  folder <- file.path(path)
    # Construct file paths for the two time periods
  file_ERA5Land <- file.path(folder, "04_result/PCWD_ANNMAX.nc")

      # Load grid data for each time period
      grid_ERA5Land <- loadGridData(dataset = file_ERA5Land, var = "pcwd_annmax")

      # Set spatial projection
      grid_ERA5Land <- setGridProj(grid = grid_ERA5Land, proj = proj4string(refregions))

      # Perform spatial overlay
      grid_region_ERA5Land <- overGrid(grid_ERA5Land, region_object)

      # Extract data arrays
      data_array_ERA5Land <- grid_region_ERA5Land$Data[]

      # Compute spatial average for each time step
      spatial_avg <- apply(data_array_ERA5Land, 1, function(slice) {
        mean(slice, na.rm = TRUE) # Compute mean for the spatial dimensions, ignoring NA
      })

      # Store the spatial average in the list
      result_array_region <- spatial_avg

  # Store the result for the current region
  regional_results_ERA5Land[[region]] <- result_array_region
}
#save calculated list
saveRDS(regional_results_ERA5Land, file="~/cwd_global/data/regionalResults_ERA5Land.RData") #ERA5Land regional averages


```

Read in regional ModESim data
```{r}
regional_results_ModESim <- readRDS("~/cwd_global/data/regionalResults_1850_1.RData")

```



Plots: 
- could show violin plots for each region, showing the values from ModESim with ERA5 overlayn to see where the match works well / or density curves
- could also show as scatterplot where data should lie on 1:1 line if it represents well 
