---
title: "Extreme Value Analysis PCWD"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(maps)
library(raster)
library(rnaturalearth)
library(sf)
```

This demonstrates the workflow for fitting an extreme value distribution to annual maxima of the CWD time series.

## Prepare data

```{r read data from netcdf, echo=FALSE}
##read in data
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850/PCWD_ANNMAX.nc"
nc_pwcd_1850 <- nc_open(input_file)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat")
time = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)

```

Select one gridcell: 
```{r}
# Specify the desired latitude and longitude (close to Munich here)
target_lon <- 11.5  # Replace with your desired longitude
target_lat <- 48.0    # Replace with your desired latitude

# Find the indices of the closest longitude and latitude
lon_index <- which.min(abs(lon - target_lon))
lat_index <- which.min(abs(lat - target_lat))

# Extract the time series for the specific grid cell
pcwd_gridcell <- pcwd_annmax_1850[lon_index, lat_index, ]

# # Now you have the time series for the specific grid cell
# print(pcwd_gridcell)

plot(time_dates, pcwd_gridcell, type = "l", xlab = "Time", ylab = "PCWD Annual Max",
     main = paste("Time Series for Grid Cell at", target_lon, target_lat))


```


## Extreme value statistics one gridcell

Get an overview plot of the data: 
```{r}
# Create a histogram
hist(pcwd_gridcell, freq = FALSE, main = "Histogram and density")

# Calculate density
dx <- density(pcwd_gridcell)

# Add density
lines(dx, lwd = 2, col = "red")

# Plot the density without histogram
plot(dx, lwd = 2, col = "red",
     main = "Density")

# Add the data-poins with noise in the X-axis
rug(jitter(pcwd_gridcell))
```

Get annual maxima and fit a general extreme value distribution using the {extRemes} package.
- GEV is used for block maxima (e.g. annual maximum PCWD)
- if the shape parameter ξ = 0, distribution follows a Gumbel distribution (see example below)
- if the shape parameter ξ > 0, distribution follows a Fréchet distribution, meaning that the extremes with higher magnitudes have a higher probability of occurrence than would be seen in a distribution where ξ = 0
- if the shape parameter ξ < 0, distribution follows a Weibull distribution
```{r}

evd_gev <- extRemes::fevd(x = pcwd_gridcell, type = "GEV", method = "MLE", units = "years")
summary(evd_gev)
```


Get CWD magnitudes for given return periods.
```{r}
return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 200, 250, 300, 500, 800)

return_level <- extRemes::return.level(
  evd_gev,
  return.period = return_period
)
df_return <- tibble(
  return_period = return_period,
  return_level = unname(c(return_level)),
  trans_period = -log( -log(1 - 1/return_period)) )

df_return |> 
  ggplot(aes(return_period, return_level)) +
  geom_point() +
  labs(x = "Return period (yr)", 
       y = "Magnitude of annual CWD maximum (mm)",
       title = "GEV")
```

With a Gumbel extreme value distribution, the return period as a function of the CWD extreme magnitude is calculated as follows:
```{r}
# Fit Gumbel distribution
evd_gumbi <- extRemes::fevd(x = pcwd_gridcell, type = "Gumbel", method = "MLE", units = "years")
summary(evd_gumbi)

# calculate return period as a function of the CWD extreme. Using the two 
# coefficients of the fitted distribution as arguments
calc_return_period <- function(x, loc, scale){
  1 / (1 - exp(-exp(-(x-loc)/scale)))
}

extract_loc <- function(mod){
  loc <- mod$results$par[ "location" ]
  if (!is.null(loc)){
    return(loc)
  } else {
    return(NA)
  }
}

extract_scale <- function(mod){
  scale <- mod$results$par[ "scale" ]
  if (!is.null(scale)){
    return(scale)
  } else {
    return(NA)
  }
}

# demo return periods
return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 200, 250, 300, 500, 800)

# use built-in function to get expected CWD extreme for given return periods 
# (inverse of probability)
return_level <- extRemes::return.level(
  evd_gumbi,
  return.period = return_period
)

# create data frame for visualisation
df_return <- tibble(
  return_period = return_period,
  return_level = unname(c(return_level)),
  trans_level = -log( -log(1 - 1/return_period))) |> 
  mutate(myreturn_period = calc_return_period(
    return_level,
    extract_loc(evd_gumbi),
    extract_scale(evd_gumbi)
  ))

# CWD extreme for a given return period
df_return |> 
  ggplot(aes(return_period, return_level)) +
  geom_point() +
  labs(x = "Return period (yr)", 
       y = "Magnitude of annual CWD maximum (mm)",
       title = "Gumbel")

# Return period for a given CWD extreme (calculated based on function above)
df_return |> 
  ggplot(aes(return_level, myreturn_period)) + 
  geom_point() +
  labs(y = "Return period (yr)", 
       x = "Magnitude of annual CWD maximum (mm)",
       title = "Gumbel")
```

Visualise the estimated event size with a return period of $T = 80$ y as the red line on top of the distribution of cumulative water deficit events.
```{r}
# Flatten the array into a vector
deficit_vector <- c(pcwd_gridcell)

ggplot() +
  geom_histogram(
    aes(x = deficit_vector, y = after_stat(density)),
    color = "black",
    position="identity",
    bins = 20
    ) +
  labs(x = "Cumulative water deficit (mm)") +
  geom_vline(xintercept = df_return %>%
               dplyr::filter(return_period == 80) %>%
               pull(return_level),
             col = "tomato")
```


Apply to all gridcells and plot: 
```{r}
#calculate return periods for each gridcell and plot 

#function that calculates EVA: 
calculate_return_level <- function(data, return_period) {
  # Fit Gumbel distribution
  evd_fit <- extRemes::fevd(x = data, type = "GEV", method = "MLE", units = "years")
  
  # Extract the return level for the specified return period
  return_level <- extRemes::return.level(evd_fit, return.period = return_period)
  
  # Return the calculated return level
  return(unname(return_level))
}

```

loop over the dataset and extract return period value: 
```{r, echo=FALSE, warnings=FALSE}
# Create an empty matrix to store 80-year return levels
return_level_80yr <- matrix(NA, nrow = length(lon), ncol = length(lat))

# Loop through each grid cell
for (i in seq_along(lon)) {
  for (j in seq_along(lat)) {
    # Extract time series for the grid cell
    data <- pcwd_annmax_1850[i, j, ]
    
    # Check for NA or insufficient data
   # if (all(is.na(data)) || length(na.omit(data)) < 10) next
    
    # Calculate 80-year return level
    return_level_80yr[i, j] <- calculate_return_level(data, return_period = 80)
  }
}

#print(return_level_80yr)
```


now plot 80 yr return values of PCWD on map: 

```{r}
# Convert first year's data into a raster object
r <- raster(t(return_level_80yr), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r<- flip(r, direction='y')
# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```

```{r}
# Define custom color bins
custom_bins <- c(0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
custom_colors <- c("#004f5a",  # New color: a darker turquoise at the beginning
                   "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", "#e8b249", 
                   "#e3842a", "#d45321", "#ba2e1f", "#9e2d2a", "#803c3d", "#63363f", 
                   "#9c276e",  # New color: a deeper shade in the pink tones
                   "#d238a5")

# Define the numeric labels for the breaks
break_labels <- c(as.character(scales::label_number(accuracy = 1)(custom_bins[-length(custom_bins)])), "Ocean")
```

```{r}
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "80 yr return level", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "PCWD ANNMAX 80 year return level m001 1850 set", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "80 yr return level", reverse = TRUE))  # Adjust legend for better presentation

ggsave("80yr_return_m001_1850_10mm.tiff", width = 20, height = 14, units = "cm", dpi=320)
```

## Now apply the extreme value statistics to IPCC regions

Load libraries from Climate4R
```{r}
library(rgdal)
library(sp)
library(RColorBrewer)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
library(geoprocessoR)
```

Load reference regions and coastlines:
```{r}
load("/storage/homefs/ph23v078/Reference_regions/IPCC-WGI-reference-regions-v4_R.rda", verbose = TRUE)

#simplify this object by converting it to a SpatialPolygons class object (i.e., only the polygons are retained and their attributes discarded):
refregions <- as(IPCC_WGI_reference_regions_v4, "SpatialPolygons")

temp.dir <- tempdir()
unzip("/storage/homefs/ph23v078/Reference_regions/ne_110m_coastline.zip", exdir = temp.dir)
coastLines <- readOGR(dsn = temp.dir, layer = "ne_110m_coastline")

```

## Selecting specific IPCC regions

For instance, in order to extract one single region (e.g., Western Central Europe, WCE):
```{r}
MED <- refregions["MED"]
NEU <- refregions["NEU"]
WCE <- refregions["WCE"]
plot(WCE)
plot(coastLines, col = "grey", add = TRUE)
text(x = coordinates(WCE)[,1],
     y = coordinates(WCE)[,2],
     labels = names(WCE), cex = 1)
```
Or all of Europe
```{r}
europe <- refregions[c("NEU", "WCE", "MED")]

plot(europe)
plot(coastLines, col = rgb(0.85,0.85,0.85,0.7), add = TRUE)
text(x = coordinates(europe)[,1],
    y = coordinates(europe)[,2],
    labels = names(europe), cex = 1)
```

## Example with ModE-Sim data:
First, dataInventory provides basic information about the variables available in the file/dataset:
```{r, message=FALSE, echo=FALSE}
di <- dataInventory("/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850_AbsTrsh/PCWD_ANNMAX.nc")
str(di)
```
 To load pcwd_annmax data, the function loadGridData from package loadeR is used:
```{r}
grid1 <- loadGridData(dataset = "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850_AbsTrsh/PCWD_ANNMAX.nc",
                      var = "pcwd_annmax")
```
## Extracting data for the region/s of interest
Function overGrid perfoms the operation of intersecting a climate4R grid and a Spatial object. The only requirement is equal projections. Use function projectGrid to define and/or change the projection of a grid. Use proj4string and spTransform for Spatial objects. In this example the map projection is WGS84 (EPSG:4326). Note that several warning messages may appear during the following operations. These arise from recent changes in the versions of PROJ and GDAL, and can be safely ignored throughout these examples.
```{r}
proj4string(refregions)
#An appropriate definition of the current coordinates projection will be later needed to ensure spatial consistency of geospatial operations:
grid1 <- setGridProj(grid = grid1, proj = proj4string(refregions))
#Once the spatial reference is defined, the spatial overlay is done:
grid1.eu <- overGrid(grid1, europe)
```


Setting the argument subset = TRUE the spatial object is cropped to the region extent:
```{r}
grid1.eu <- overGrid(grid1, europe, subset = TRUE)
```

```{r}
regnameslayer <- list("sp.text", coordinates(refregions), names(refregions))
spatialPlot(climatology(grid1.eu), backdrop.theme = "coastline",
            color.theme = "BuGn",
            rev.colors = FALSE,
            sp.layout = list(list(refregions, first = FALSE, col = "blue"), regnameslayer))


```
Now apply extreme value statistics on the region of Europe:

Get an overview plot of the data: 
```{r}
# Create a histogram
hist(grid1.eu$Data, freq = FALSE, main = "Histogram and density")

# Calculate density
dx <- density(grid1.eu$Data)

# Add density
lines(dx, lwd = 2, col = "red")

# Plot the density without histogram
plot(dx, lwd = 2, col = "red",
     main = "Density")

# Add the data-poins with noise in the X-axis
rug(jitter(grid1.eu$Data))
```
comments: still have northern Africa contained in the MED area so exclude that from the Europe cell? Or compute for the smaller areas? 
e.g. just for CWE: 
```{r}
proj4string(refregions)
grid1.wce <- overGrid(grid1, WCE)
grid1.wce <- overGrid(grid1, WCE, subset = TRUE)

#show data: 
# Create a histogram
hist(grid1.wce$Data, freq = FALSE, main = "Histogram and density")

# Calculate density
dx <- density(grid1.wce$Data, na.rm=TRUE)

# Add density
lines(dx, lwd = 2, col = "red")

# Plot the density without histogram
plot(dx, lwd = 2, col = "red",
     main = "Density")

# Add the data-poins with noise in the X-axis
rug(jitter(grid1.wce$Data))
```

```{r}
grid185001 <- subsetGrid(grid1.wce, years = 1850, drop = TRUE)

regnameslayer <- list("sp.text", coordinates(refregions), names(refregions))
spatialPlot(grid185001, backdrop.theme = "coastline",
            color.theme = "BuGn",
            rev.colors = FALSE,
            sp.layout = list(list(refregions, first = FALSE, col = "blue"), regnameslayer))


```
Now calculate the EVS for this area: 
```{r}
grid1.wce_EVA <- as.numeric(grid1.wce$Data)
grid1.wce_EVA <- na.omit(grid1.wce_EVA)
evd_gev.wce <- extRemes::fevd(x = grid1.wce_EVA, type = "GEV", method = "MLE", units = "years")
summary(evd_gev.wce)
```
```{r}
return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 200, 250, 300, 500, 800)

return_level.wce <- extRemes::return.level(
  evd_gev.wce,
  return.period = return_period
)
df_return <- tibble(
  return_period = return_period,
  return_level = unname(c(return_level.wce)),
  trans_period = -log( -log(1 - 1/return_period)) )

df_return |> 
  ggplot(aes(return_period, return_level)) +
  geom_point() +
  labs(x = "Return period (yr)", 
       y = "Magnitude of annual CWD maximum (mm)",
       title = "GEV - 1850 WCE")
```
Visualise the estimated event size with a return period of $T = 80$ y as the red line on top of the distribution of cumulative water deficit events.
```{r}
# Flatten the array into a vector
deficit_vector <- c(grid1.wce_EVA)

ggplot() +
  geom_histogram(
    aes(x = deficit_vector, y = after_stat(density)),
    color = "black",
    position="identity",
    bins = 20
    ) +
  labs(x = "Cumulative water deficit (mm)") +
  geom_vline(xintercept = df_return %>%
               dplyr::filter(return_period == 80) %>%
               pull(return_level),
             col = "tomato")
```

loop over the dataset and extract return period value: 
```{r, echo=FALSE, warnings=FALSE}
# Extract dimensions and prepare storage
x_coords <- grid1.wce$xyCoords$x
y_coords <- grid1.wce$xyCoords$y

# Dimensions of the Data array
num_x <- length(x_coords)
num_y <- length(y_coords)

# Create an empty matrix to store 80-year return levels
return_level_80yr.wce <- matrix(NA, nrow = num_x, ncol = num_y)

# Loop through each grid cell
for (i in seq_len(num_x)) {
  for (j in seq_len(num_y)) {
    # Extract time series for the grid cell
    data <- grid1.wce$Data[, j, i]  # Adjust indexing: time (1st dim), y (2nd dim), x (3rd dim)
    
    # Check for NA or insufficient data
    if (all(is.na(data)) || length(na.omit(data)) < 10) next
    
    # Calculate 80-year return level
    return_level_80yr.wce[i, j] <- calculate_return_level(data, return_period = 80)
  }
}

```

```{r}
#IPCC region inserts
# Convert WCE to an sf object
WCE_sf <- st_as_sf(WCE)
# Ensure the CRS matches with the raster and other layers
WCE_sf <- st_transform(WCE_sf, crs = st_crs(r.wce))
# Convert MED and NEU to sf objects
MED_sf <- st_as_sf(MED)
NEU_sf <- st_as_sf(NEU)
# Ensure the CRS matches with the raster and other layers
MED_sf <- st_transform(MED_sf, crs = st_crs(r.wce))
NEU_sf <- st_transform(NEU_sf, crs = st_crs(r.wce))

# Calculate centroids for text placement
WCE_centroids <- st_centroid(WCE_sf)
# Add a column for region names (if not already available)
WCE_centroids$region_name <- "WCE"  # Replace "WCE" with the actual region name if available
# Calculate centroids for text placement
NEU_centroids <- st_centroid(NEU_sf)
# Add a column for region names (if not already available)
NEU_centroids$region_name <- "NEU"  # Replace "WCE" with the actual region name if available
# Calculate centroids for text placement
MED_centroids <- st_centroid(MED_sf)
# Add a column for region names (if not already available)
MED_centroids$region_name <- "MED"  # Replace "WCE" with the actual region name if available

# Coordinates (longitude and latitude)
lon.eu <- grid1.eu$xyCoords$x
lat.eu <- grid1.eu$xyCoords$y

```


```{r}
# Coordinates (longitude and latitude)
lon.wce <- grid1.wce$xyCoords$x
lat.wce <- grid1.wce$xyCoords$y

# Convert into a raster object
r.wce <- raster(t(return_level_80yr.wce), xmn=min(lon.wce), xmx=max(lon.wce), ymn=min(lat.wce), ymx=max(lat.wce))
crs(r.wce) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r.wce<- flip(r.wce, direction='y')
# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r.wce))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked.wce <- mask(r.wce, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df.wce <- as.data.frame(r_masked.wce, xy = TRUE)
colnames(r_df.wce) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom bins based on the actual data range
custom_bins <- c(200, 300, 400, 500, 600, 700, 800, 900)  # Adjust bins for your data range

# Update custom color palette to match the number of bins
custom_colors <- c("#004f5a", "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", 
                   "#e8b249", "#e3842a", "#d45321", "#ba2e1f", "#9e2d2a")

# Create labels for the legend
break_labels <- paste0(custom_bins[-length(custom_bins)], " - ", custom_bins[-1])

# Plot with ggplot2
ggplot() +
  # Raster layer for return levels
  geom_raster(data = r_df.wce, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  # Custom color palette for return levels
  scale_fill_manual(
    name = "80 yr return level", 
    values = custom_colors, 
    na.value = "lightgrey", 
    labels = break_labels,
    drop = FALSE
  ) +
  # Add title and axis labels
  labs(
    title = "PCWD ANNMAX 80 year return level m001 1850 set",
    x = "Longitude",
    y = "Latitude"
  ) +
  # Base theme and styling
  theme_classic() +
  # Overlay land boundaries
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  # Overlay WCE outline
  geom_sf(data = WCE_sf, fill = NA, color = "blue", lwd = 0.8) +
  geom_sf(data = NEU_sf, fill = NA, color = "blue", lwd = 0.8) +
  geom_sf(data = MED_sf, fill = NA, color = "blue", lwd = 0.8) +
  # Add region name labels
  geom_text(
    data = WCE_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = WCE_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  geom_text(
    data = NEU_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = NEU_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  geom_text(
    data = MED_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = MED_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  # Ensure proper coordinate scaling
  coord_sf(xlim = range(lon.eu), ylim = range(lat.eu), expand = FALSE) +
  # Adjust legend styling
  theme(
    legend.title = element_text(size = 12),  # Adjust legend title size
    legend.text = element_text(size = 10)    # Adjust legend text size
  ) +
  # Adjust legend presentation
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE, override.aes = list(color = NA)))

```

### Comparison between 1420 and 1850 dataset

```{r}
grid2 <- loadGridData(dataset = "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1420_AbsTrsh/PCWD_ANNMAX.nc",
                      var = "pcwd_annmax")

proj4string(refregions)
#An appropriate definition of the current coordinates projection will be later needed to ensure spatial consistency of geospatial operations:
grid2 <- setGridProj(grid = grid2, proj = proj4string(refregions))
#Once the spatial reference is defined, the spatial overlay is done:
grid2.wce <- overGrid(grid2, WCE, subset = TRUE)

#show data: 
# Create a histogram
hist(grid2.wce$Data, freq = FALSE, main = "Histogram and density")

# Calculate density
dx2 <- density(grid2.wce$Data, na.rm=TRUE)

# Add density
lines(dx2, lwd = 2, col = "red")

# Plot the density without histogram
plot(dx2, lwd = 2, col = "red",
     main = "Density")

# Add the data-poins with noise in the X-axis
rug(jitter(grid2.wce$Data))

```
```{r}
# Plot the density without histogram
plot(dx, lwd = 2, col = "red",
     main = "Density plots of m001 WCE annmax PCWD", ylim=c(0, 0.0030))
# Add density for 1850 in blue
lines(dx2, lwd = 2, col = "blue")
# Add a legend
legend("topright", inset=.02, legend=c("1850", "1420"),
       col=c("red", "blue"), lty=1, cex=0.8)


```
Now calculate the EVS for this area: 
```{r}
grid2.wce_EVA <- as.numeric(grid2.wce$Data)
grid2.wce_EVA <- na.omit(grid2.wce_EVA)
evd_gev2.wce <- extRemes::fevd(x = grid2.wce_EVA, type = "GEV", method = "MLE", units = "years")
summary(evd_gev2.wce)
```
```{r}
return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 120, 200, 250, 300, 500, 800)

return_level2.wce <- extRemes::return.level(
  evd_gev2.wce,
  return.period = return_period
)
df_return2 <- tibble(
  return_period = return_period,
  return_level = unname(c(return_level2.wce)),
  trans_period = -log( -log(1 - 1/return_period)) )

df_return2 |> 
  ggplot(aes(return_period, return_level)) +
  geom_point() +
  labs(x = "Return period (yr)", 
       y = "Magnitude of annual CWD maximum (mm)",
       title = "GEV - WCE 1420")
```

```{r}
# Flatten the array into a vector
deficit_vector2 <- c(grid2.wce_EVA)

ggplot() +
  scale_color_manual(name = "sets", 
                     values = c("1420" = "blue", "1850" = "tomato")) +
  geom_histogram(
    aes(x = deficit_vector2, y = after_stat(density),
    color = "1420"),
    position="identity",
    bins = 20, alpha=I(0.5)
    ) +
  geom_histogram(
    aes(x = deficit_vector, y = after_stat(density),
    color = "1850"),
    position="identity",
    bins = 20, alpha=I(0.5)
    ) +
  labs(x = "Cumulative water deficit (mm)") +
  geom_vline(xintercept = df_return2 %>%
               dplyr::filter(return_period == 80) %>%
               pull(return_level),
             col = "blue")+
  geom_vline(xintercept = df_return %>%
               dplyr::filter(return_period == 80) %>%
               pull(return_level),
             col = "tomato")+
  theme_classic2()
```
loop over the dataset and extract return period value: 
```{r, echo=FALSE, warnings=FALSE}
# Extract dimensions and prepare storage
x_coords <- grid2.wce$xyCoords$x
y_coords <- grid2.wce$xyCoords$y

# Dimensions of the Data array
num_x <- length(x_coords)
num_y <- length(y_coords)

# Create an empty matrix to store 80-year return levels
return_level_80yr2.wce <- matrix(NA, nrow = num_x, ncol = num_y)

# Loop through each grid cell
for (i in seq_len(num_x)) {
  for (j in seq_len(num_y)) {
    # Extract time series for the grid cell
    data <- grid2.wce$Data[, j, i]  # Adjust indexing: time (1st dim), y (2nd dim), x (3rd dim)
    
    # Check for NA or insufficient data
    if (all(is.na(data)) || length(na.omit(data)) < 10) next
    
    # Calculate 80-year return level
    return_level_80yr2.wce[i, j] <- calculate_return_level(data, return_period = 80)
  }
}

```
```{r}
# Coordinates (longitude and latitude)
lon.wce <- grid2.wce$xyCoords$x
lat.wce <- grid2.wce$xyCoords$y

# Convert first year's data into a raster object
r2.wce <- raster(t(return_level_80yr2.wce), xmn=min(lon.wce), xmx=max(lon.wce), ymn=min(lat.wce), ymx=max(lat.wce))
crs(r2.wce) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r2.wce<- flip(r2.wce, direction='y')
# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r2.wce))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked2.wce <- mask(r2.wce, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df2.wce <- as.data.frame(r_masked2.wce, xy = TRUE)
colnames(r_df2.wce) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom bins based on the actual data range
custom_bins <- c(200, 300, 400, 500, 600, 700, 800, 900)  # Adjust bins for your data range

# Update custom color palette to match the number of bins
custom_colors <- c("#004f5a", "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", 
                   "#e8b249", "#e3842a", "#d45321", "#ba2e1f", "#9e2d2a")
# Create labels for the legend
break_labels <- paste0(custom_bins[-length(custom_bins)], " - ", custom_bins[-1])

# Plot with ggplot2
ggplot() +
  # Raster layer for return levels
  geom_raster(data = r_df2.wce, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  # Custom color palette for return levels
  scale_fill_manual(
    name = "80 yr return level", 
    values = custom_colors, 
    na.value = "lightgrey", 
    labels = break_labels, 
    drop = FALSE  # Ensures only defined bins are shown
  ) +
  # Add title and axis labels
  labs(
    title = "PCWD ANNMAX 80 year return level m001 1420 set",
    x = "Longitude",
    y = "Latitude"
  ) +
  # Base theme and styling
  theme_classic() +
  # Overlay land boundaries
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  # Overlay WCE outline
  geom_sf(data = WCE_sf, fill = NA, color = "blue", lwd = 0.8) +
  geom_sf(data = NEU_sf, fill = NA, color = "blue", lwd = 0.8) +
  geom_sf(data = MED_sf, fill = NA, color = "blue", lwd = 0.8) +
  # Add region name labels
  geom_text(
    data = WCE_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = WCE_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  geom_text(
    data = NEU_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = NEU_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  geom_text(
    data = MED_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = MED_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  # Ensure proper coordinate scaling
  coord_sf(xlim = range(lon.eu), ylim = range(lat.eu), expand = FALSE) +
  # Adjust legend styling
  theme(
    legend.title = element_text(size = 12),  # Adjust legend title size
    legend.text = element_text(size = 10)    # Adjust legend text size
  ) +
  # Adjust legend presentation
  guides(fill = guide_legend(title = "PWCD [mm]", reverse = TRUE, override.aes = list(color = NA)))

```

Difference between both (1850-1420): 
```{r}
return_level_diff.wce = return_level_80yr.wce - return_level_80yr2.wce
```


```{r}
# Coordinates (longitude and latitude)
lon.wce <- grid1.wce$xyCoords$x
lat.wce <- grid1.wce$xyCoords$y

# Convert first year's data into a raster object
r.wce <- raster(t(return_level_diff.wce), xmn=min(lon.wce), xmx=max(lon.wce), ymn=min(lat.wce), ymx=max(lat.wce))
crs(r.wce) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r.wce<- flip(r.wce, direction='y')
# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r.wce))  # Ensure CRS alignment
# Convert WCE to an sf object
WCE_sf <- st_as_sf(WCE)
# Ensure the CRS matches with the raster and other layers
WCE_sf <- st_transform(WCE_sf, crs = st_crs(r.wce))
# Calculate centroids for text placement
WCE_centroids <- st_centroid(WCE_sf)
# Add a column for region names (if not already available)
WCE_centroids$region_name <- "WCE"  # Replace "WCE" with the actual region name if available


# Mask the raster with land polygons to remove ocean values
r_masked.wce <- mask(r.wce, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df.wce <- as.data.frame(r_masked.wce, xy = TRUE)
colnames(r_df.wce) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

custom_bins <- c(-300,-100, -50, -20, -10, -1, 0, 1, 10, 20, 50, 100, Inf)  # Define breaks based on data range
# Create a 10-color scale from blue to white to red
custom_colors <- colorRampPalette(c("blue", "white", "red"))(13)
# Create labels for the legend
break_labels <- paste0(custom_bins[-length(custom_bins)], " - ", custom_bins[-1])

# Plot with ggplot2
ggplot() +
  # Raster layer for return levels
  geom_raster(data = r_df.wce, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  # Custom color palette for return levels
  scale_fill_manual(
    name = "80 yr return level", 
    values = custom_colors, 
    na.value = "lightgrey", 
    labels = break_labels,
    drop = FALSE
  ) +
  # Add title and axis labels
  labs(
    title = "PCWD ANNMAX 80 year return level change 1850-1420",
    x = "Longitude",
    y = "Latitude"
  ) +
  # Base theme and styling
  theme_classic() +
  # Overlay land boundaries
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  # Overlay WCE outline
  geom_sf(data = WCE_sf, fill = NA, color = "blue", lwd = 0.8) +
  geom_sf(data = NEU_sf, fill = NA, color = "blue", lwd = 0.8) +
  geom_sf(data = MED_sf, fill = NA, color = "blue", lwd = 0.8) +
  # Add region name labels
  geom_text(
    data = WCE_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = WCE_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  geom_text(
    data = NEU_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = NEU_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  geom_text(
    data = MED_centroids %>% st_coordinates() %>% as_tibble() %>% 
      mutate(region_name = MED_centroids$region_name), 
    aes(x = X, y = Y, label = region_name), 
    color = "blue", 
    size = 4
  ) +
  # Ensure proper coordinate scaling
  coord_sf(xlim = range(lon.eu), ylim = range(lat.eu), expand = FALSE) +
  # Adjust legend styling
  theme(
    legend.title = element_text(size = 12),  # Adjust legend title size
    legend.text = element_text(size = 10)    # Adjust legend text size
  ) +
  # Adjust legend presentation
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE, override.aes = list(color = NA)))

```
## Analysis on all ensemble members - first Model Set 

Aggregate all 1850 files: 

```{r}
### append data i.e. get rid of time dimension and just append to exisiting data
# Load necessary libraries
library(ncdf4)

# Define the base path
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Identify the target files
target_files <- file.path(folders, "04_result_1850/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included


# Initialize an array to store the extended data
# Extract dimensions from the first file to initialize the array
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Prepare an array to store the extended data
extended_data <- array(NA, dim = c(dim_x, dim_y, dim_z * length(target_files)))

# Loop through all the target files to extract the data and append along the third dimension
current_index <- 1
for (file in target_files) {
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Determine the slice of the extended array to update
  slice_indices <- current_index:(current_index + dim(data)[3] - 1)
  
  # Append the data to the extended array
  extended_data[,,slice_indices] <- data
  
  # Update the current index
  current_index <- current_index + dim(data)[3]
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the extended data
print(dim(extended_data))  # Check dimensions (x, y, extended z)

# Optionally, inspect a slice of the extended data for verification
print(extended_data[,,1:10])  # Inspect the first 10 slices
```

```{r}
#do EVA for all this
#use function defined above
# Create an empty matrix to store 80-year return levels
return_level_80yr_13EMs <- matrix(NA, nrow = length(lon), ncol = length(lat))

# Loop through each grid cell
for (i in seq_along(lon)) {
  for (j in seq_along(lat)) {
    # Extract time series for the grid cell
    data <- extended_data[i, j, ]
    
    # Check for NA or insufficient data
   # if (all(is.na(data)) || length(na.omit(data)) < 10) next
    
    # Calculate 80-year return level
    return_level_80yr_13EMs[i, j] <- calculate_return_level(data, return_period = 80)
  }
}

```

```{r}
# Convert first year's data into a raster object
r <- raster(t(return_level_80yr_13EMs), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r<- flip(r, direction='y')
# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations
```

```{r}
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "80 yr return level", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "PCWD ANNMAX 80-yr RL m001-m013 1850 set", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "80 yr return level", reverse = TRUE))  # Adjust legend for better presentation

ggsave("80yr_return_m001-m013_1850_10mm.tiff", width = 20, height = 14, units = "cm", dpi=320)
```


