---
title: "Results"
author: "Patricia Gribi"
date: "2024-07-24"
output: html_document
---

von 5 Hauptklimazonen

```{r}

# loading libraries
library(ggplot2)
library(dplyr)
library(terra)
library(tidyr)


# read rds files

## annmax files
cwd_df <- readr::read_rds("/data_1/CMIP6/tidy/cwd/cwd_ANNMAX.rds")
pcwd_df <- readr::read_rds("/data_1/CMIP6/tidy/pcwd/pcwd_ANNMAX.rds")


cwd_df <- cwd_df |>
  select(-ilon)|>
  rename(max_deficit_cwd = "max_deficit")

pcwd_df <- pcwd_df |>
  select(-ilon)|>
  rename(max_deficit_pcwd = "max_deficit")


combined_annmax_df <- cwd_df |>  # one of the daily data frames
    left_join(pcwd_df, by = join_by(lon, lat, year))


```


# Desert

## Annual Mean CWD & PCWD Time Series

```{r}

source(paste0(here::here(), "/R/get_coords.R"))

combined_df <- get_coords(4, 28)

combined_mean_df <- combined_df |>
      mutate(year = lubridate::year(time)) |>
      group_by(year) |>
      summarise(mean_cwd_deficit = mean(deficit_cwd), mean_pcwd_deficit = mean(deficit_pcwd))

ggplot(combined_mean_df, aes(x = year)) +
  geom_line(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_line(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_point(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_point(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_smooth(aes(y = mean_cwd_deficit, color = "CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  geom_smooth(aes(y = mean_pcwd_deficit, color = "PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  labs(title = "Annual Mean CWD and PCWD Values",
       subtitle = "Location: Algeria (Longitude 3.75, Latitude 28.7)",
       x = "Year",
       y = "Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```

## Annual Max CWD & PCWD Time Series

```{r}

# extract longitude
lon <- 3.75

combined_annmax_df <- combined_annmax_df[combined_annmax_df$lon == lon, ]

# extract nearest latitude 50.4 
start <- 28
end <- 29
combined_annmax_df <- combined_annmax_df[(combined_annmax_df$lat >= start) & (combined_annmax_df$lat <= end), ]


ggplot(combined_annmax_df, aes(x = year)) +
  geom_line(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add a line for max_deficit_cwd
  geom_line(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add a line for max_deficit_pcwd
  geom_point(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add points for max_deficit_cwd
  geom_point(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add points for max_deficit_pcwd
  geom_smooth(aes(y = max_deficit_cwd, color = "Max CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_cwd
  geom_smooth(aes(y = max_deficit_pcwd, color = "Max PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_pcwd
  labs(title = "Annual Max CWD and PCWD Values",
       subtitle = "Location: Algeria (Longitude 3.75, Latitude 28.7)",
       x = "Year",
       y = "Max Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```

Möglicherweise stimmt das schon denn in der Sahara ist das CWD aufgrund fehlenden Wassers limitiert. Und wäre PCWD nicht, wenn es nicht wasser limitiert wäre?



# Tropics

## Annual Mean CWD & PCWD Time Series

```{r}

source(paste0(here::here(), "/R/get_coords.R"))

combined_df <- get_coords(247, -8)

combined_mean_df <- combined_df |>
      mutate(year = lubridate::year(time)) |>
      group_by(year) |>
      summarise(mean_cwd_deficit = mean(deficit_cwd), mean_pcwd_deficit = mean(deficit_pcwd))

ggplot(combined_mean_df, aes(x = year)) +
  geom_line(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_line(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_point(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_point(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_smooth(aes(y = mean_cwd_deficit, color = "CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  geom_smooth(aes(y = mean_pcwd_deficit, color = "PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  labs(title = "Annual Mean CWD and PCWD Values",
       subtitle = "Location: Brasil (Longitude 307.5, Latitude -7)",
       x = "Year",
       y = "Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```

## Annual Max CWD & PCWD Time Series

```{r}

# extract longitude
lon <- 307.5

combined_annmax_df <- combined_annmax_df[combined_annmax_df$lon == lon, ]

# extract nearest latitude 
start <- -9
end <- -8
combined_annmax_df <- combined_annmax_df[(combined_annmax_df$lat >= start) & (combined_annmax_df$lat <= end), ]


ggplot(combined_annmax_df, aes(x = year)) +
  geom_line(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add a line for max_deficit_cwd
  geom_line(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add a line for max_deficit_pcwd
  geom_point(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add points for max_deficit_cwd
  geom_point(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add points for max_deficit_pcwd
  geom_smooth(aes(y = max_deficit_cwd, color = "Max CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_cwd
  geom_smooth(aes(y = max_deficit_pcwd, color = "Max PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_pcwd
  labs(title = "Annual Max CWD and PCWD Values",
       subtitle = "Location: Brasil (Longitude 307.5, Latitude -7)",
       x = "Year",
       y = "Max Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```

das pcwd kann hier doch gar nicht sein... warum so hohe Werte?



# Ice 

## Annual Mean CWD & PCWD Time Series

```{r}

source(paste0(here::here(), "/R/get_coords.R"))

combined_df <- get_coords(45, -81)

combined_mean_df <- combined_df |>
      mutate(year = lubridate::year(time)) |>
      group_by(year) |>
      summarise(mean_cwd_deficit = mean(deficit_cwd), mean_pcwd_deficit = mean(deficit_pcwd))

ggplot(combined_mean_df, aes(x = year)) +
  geom_line(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_line(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_point(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_point(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_smooth(aes(y = mean_cwd_deficit, color = "CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  geom_smooth(aes(y = mean_pcwd_deficit, color = "PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  labs(title = "Annual Mean CWD and PCWD Values",
       subtitle = "Location: Antarctica (Longitude 55, Latitude -80.6)",
       x = "Year",
       y = "Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```

## Annual Max CWD & PCWD Time Series

```{r}

# extract longitude
lon <- 55

combined_annmax_df <- combined_annmax_df[combined_annmax_df$lon == lon, ]

# extract nearest latitude 
start <- -81
end <- -80
combined_annmax_df <- combined_annmax_df[(combined_annmax_df$lat >= start) & (combined_annmax_df$lat <= end), ]


ggplot(combined_annmax_df, aes(x = year)) +
  geom_line(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add a line for max_deficit_cwd
  geom_line(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add a line for max_deficit_pcwd
  geom_point(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add points for max_deficit_cwd
  geom_point(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add points for max_deficit_pcwd
  geom_smooth(aes(y = max_deficit_cwd, color = "Max CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_cwd
  geom_smooth(aes(y = max_deficit_pcwd, color = "Max PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_pcwd
  labs(title = "Annual Max CWD and PCWD Values",
       subtitle = "Location: Antarctica (Longitude 55, Latitude -80.6)",
       x = "Year",
       y = "Max Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```



# Tundra

## Annual Mean CWD & PCWD Time Series

```{r}

source(paste0(here::here(), "/R/get_coords.R"))

combined_df <- get_coords(90, 71)

combined_mean_df <- combined_df |>
      mutate(year = lubridate::year(time)) |>
      group_by(year) |>
      summarise(mean_cwd_deficit = mean(deficit_cwd), mean_pcwd_deficit = mean(deficit_pcwd))

ggplot(combined_mean_df, aes(x = year)) +
  geom_line(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_line(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_point(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_point(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_smooth(aes(y = mean_cwd_deficit, color = "CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  geom_smooth(aes(y = mean_pcwd_deficit, color = "PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  labs(title = "Annual Mean CWD and PCWD Values",
       subtitle = "Location: Russia (Longitude 111.25, Latitude 71.2)",
       x = "Year",
       y = "Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```

## Annual Max CWD & PCWD Time Series

```{r}

# extract longitude
lon <- 111.25

combined_annmax_df <- combined_annmax_df[combined_annmax_df$lon == lon, ]

# extract nearest latitude 
start <- 71
end <- 72
combined_annmax_df <- combined_annmax_df[(combined_annmax_df$lat >= start) & (combined_annmax_df$lat <= end), ]


ggplot(combined_annmax_df, aes(x = year)) +
  geom_line(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add a line for max_deficit_cwd
  geom_line(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add a line for max_deficit_pcwd
  geom_point(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add points for max_deficit_cwd
  geom_point(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add points for max_deficit_pcwd
  geom_smooth(aes(y = max_deficit_cwd, color = "Max CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_cwd
  geom_smooth(aes(y = max_deficit_pcwd, color = "Max PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_pcwd
  labs(title = "Annual Max CWD and PCWD Values",
       subtitle = "Location: Russia (Longitude 111.25, Latitude 71.2)",
       x = "Year",
       y = "Max Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```


# Temperate Forest

Czech Republic, 
selected because very near to Tharandt site of fluxnet data

ilon 11 * 1.25 = 13.75
latitude 50.41885

## Annual Mean CWD & PCWD Time Series

```{r}

source(paste0(here::here(), "/R/get_coords.R"))

combined_df <- get_coords(12, 50)

combined_mean_df <- combined_df |>
      mutate(year = lubridate::year(time)) |>
      group_by(year) |>
      summarise(mean_cwd_deficit = max(deficit_cwd), mean_pcwd_deficit = mean(deficit_pcwd))

ggplot(combined_mean_df, aes(x = year)) +
  geom_line(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_line(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_point(aes(y = mean_cwd_deficit, color = "CWD"), size = 0.5) +  
  geom_point(aes(y = mean_pcwd_deficit, color = "PCWD"), size = 0.5) + 
  geom_smooth(aes(y = mean_cwd_deficit, color = "CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  geom_smooth(aes(y = mean_pcwd_deficit, color = "PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + 
  labs(title = "Annual Mean CWD and PCWD Values",
       subtitle = "Location: Czech Republic (Longitude 13.75, Latitude 50.4)",
       x = "Year",
       y = "Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```



## Annual Max CWD & PCWD Time Series

```{r}

# extract longitude
lon <- 13.75

combined_annmax_df <- combined_annmax_df[combined_annmax_df$lon == lon, ]

# extract nearest latitude 50.4 
start <- 50
end <- 51
combined_annmax_df <- combined_annmax_df[(combined_annmax_df$lat >= start) & (combined_annmax_df$lat <= end), ]


ggplot(combined_annmax_df, aes(x = year)) +
  geom_line(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add a line for max_deficit_cwd
  geom_line(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add a line for max_deficit_pcwd
  geom_point(aes(y = max_deficit_cwd, color = "Max CWD")) +  # Add points for max_deficit_cwd
  geom_point(aes(y = max_deficit_pcwd, color = "Max PCWD")) + # Add points for max_deficit_pcwd
  geom_smooth(aes(y = max_deficit_cwd, color = "Max CWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_cwd
  geom_smooth(aes(y = max_deficit_pcwd, color = "Max PCWD"), method = "lm", se = FALSE, linetype = "dashed", size = 0.5) + # Add trendline for max_deficit_pcwd
  labs(title = "Annual Max CWD and PCWD Values",
       subtitle = "Location: Czech Republic (Longitude 13.75, Latitude 50.4)",
       x = "Year",
       y = "Max Deficit (mm/day)",
       color = "Legend") +
  theme_minimal() +    # Use a minimal theme for a clean look
  theme(axis.title.x = element_text(margin = margin(t = 15)),
        axis.title.y = element_text(margin = margin(r = 15)))

```











