---
title: "test_pcwd_US-Ton"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
library(ncdf4)
library(readr)
library(reshape2)
```

#The next section opens an annual max netcdf file which only contains the longitude of the US-Ton site: 


```{r read data from netcdf}
##read in data
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1420/PCWD_ANNMAX.nc"
nc_pwcd_1420 <- nc_open(input_file)
pcwd_annmax_1420 = ncvar_get(nc_pwcd_1420, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1420, varid="lon")
lat = ncvar_get(nc_pwcd_1420, varid="lat")
lon = ncvar_get(nc_pwcd_1420, varid="lon")
time = ncvar_get(nc_pwcd_1420, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1420)

```

```{r plot annmax}
plot(time_dates, pcwd_annmax_1420[111.25,71.2,1:431],type="l")

```

#Now read in the intermediate rds file to get information on cwd events
```{r read in RDS}
filename <- "~/scratch2/tidy/02_pcwd/ModESim_pcwd_LON_-120.000.rds"
pcwd <- read_rds(filename)
pcwd_deficit <- pcwd |> tidyr::unnest(data)

# Filter for the specific latitude and longitude (US-Ton site)
pcwd_deficit_fil <- pcwd_deficit |> 
  filter(dplyr::near(lat, 38.237736) & dplyr::near(lon, -120.000))  # Adjust lon if needed

# Select only the 'deficit' and 'date' columns
pcwd_deficit_fil <- pcwd_deficit_fil |> select(date, deficit)

# View the filtered data
print(pcwd_deficit_fil)
```


```{r plot seasonal cycle PCWD}
#select 10 years to plot
pcwd_deficit_fil_10 <- pcwd_deficit_fil |> filter(date >= as.Date("1420-01-01") & date <= as.Date("1430-01-01"))

#plots seasonal cycle of PCWD 
ggplot() +
  geom_rect(
    data = pcwd_deficit_fil_10,
    aes(xmin = min(pcwd_deficit_fil_10$date), xmax = max(pcwd_deficit_fil_10$date) , ymin = 0, ymax = max(pcwd_deficit_fil_10$deficit)),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  pcwd_deficit_fil, aes(date, deficit), color = "tomato") +
  theme_classic() +
  ylim(0, max(pcwd_deficit_fil_10$deficit)) +
  xlim(min(pcwd_deficit_fil_10$date), max(pcwd_deficit_fil_10$date))+
  labs(
    x = "Date", 
    y = "Potential cumulative water deficit (mm)"
    )
```


##-----------------------------------------------------------------------------
#Read in a test cwd file to explore analysis options

```{r read in test RDS}
filename <- "~/ModESim/tidy/02_pcwd/ModESim_pcwd_LON_-120.000.rds"
pcwd <- read_rds(filename)
pcwd_deficit <- pcwd |> tidyr::unnest(data)

# Filter for the specific latitude and longitude (US-Ton site)
pcwd_deficit_fil <- pcwd_deficit |> 
  filter(dplyr::near(lat, 38.237736) & dplyr::near(lon, -120.000))  # Adjust lon if needed

# Select only the 'deficit' and 'date' columns
#pcwd_deficit_fil <- pcwd_deficit_fil |> select(date, deficit)

# View the filtered data
print(pcwd_deficit_fil)
```

Plot the potential cumulative water deficit time series and events.
```{r}

# Filtering the dates
# Set the date range for filtering
start_date <- as.Date("1420-01-01")
end_date <- as.Date("1430-01-01")

# Filter the df tibble for dates in the specified range
filtered_df <- pcwd_deficit_fil$data[[1]]$df |> 
  filter(date >= start_date & date <= end_date)

# Filter the inst tibble based on date_start and date_end
filtered_inst <- pcwd_deficit_fil$data[[1]]$inst |> 
  filter(date_start <= end_date & date_end >= start_date)

# Combine the filtered df and inst back into the original tibble, keeping structure
pcwd_deficit_fil_filtered <- pcwd_deficit_fil |> 
  mutate(data = list(list(inst = filtered_inst, df = filtered_df)))

# Print the filtered structure
print(pcwd_deficit_fil_filtered)

ggplot() +
  geom_rect(
    data = pcwd_deficit_fil$data$inst,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = max(pcwd_deficit_fil$data$df$deficit)),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  pcwd_deficit_fil$data$df, aes(date, deficit), color = "tomato") +
  theme_classic() +
  ylim(0, max(pcwd_deficit_fil$data$df$deficit)) +
  labs(
    x = "Date", 
    y = "Potential cumulative water deficit (mm)"
    )
```
##comparing the 1850 and 1420 model sets 
- Note from Ralf: should treat the year 1850 as part of the 1850 set, not 1420
```{r read in data both sets}
#read in dataframes
##read in data from m001 1420 set
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1420/PCWD_ANNMAX.nc"
nc_pwcd_1420 <- nc_open(input_file)
pcwd_annmax_1420 = ncvar_get(nc_pwcd_1420, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1420, varid="lon")
lat = ncvar_get(nc_pwcd_1420, varid="lat")time = ncvar_get(nc_pwcd_1420, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1420)
# Assuming 'pcwd_annmax_1420' is a 3D array with dimensions (lat, lon, time)
# Convert to a dataframe using melt
melt_pcwd_annmax_1420 <- melt(pcwd_annmax_1420, varnames = c("lat", "lon", "time"), value.name = "pcwd_annmax")


##read in data from m001 1850 set
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850/PCWD_ANNMAX.nc"
nc_pwcd_1850 <- nc_open(input_file)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
#lon = ncvar_get(nc_pwcd_1850, varid="lon")
#lat = ncvar_get(nc_pwcd_1850, varid="lat")
time_1850 = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates_1850 <- reference_date + time_1850

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)
melt_pcwd_annmax_1850 <- melt(pcwd_annmax_1850, varnames = c("lat", "lon", "time"), value.name = "pcwd_annmax")

```


```{r plotting several annmax data}
#used stations used in Patricia Gribis Bachelor thesis 
####################Russian site


time <- c(1420:2009)
# Find the index of the closest latitude to 71.2
lat_index <- which.min(abs(as.numeric(lat) - 71.2))
lon_index <- which.min(abs(as.numeric(lon) - 111.25))
Russia_1420 <- pcwd_annmax_1420[lon_index,lat_index,1:431]
Russia_1850 <- pcwd_annmax_1850[lon_index,lat_index,1:160]
# Determine the maximum length between the two lists
max_length <- max(length(time), length(Russia_1420))
length(Russia_1420) <- max_length
prepend_list2 <- max_length - length(Russia_1850)
# Prepend NAs to the shorter list
Russia_1850 <- c(rep(NA, prepend_list2), Russia_1850)
Russia_df <- data.frame(time, Russia_1420, Russia_1850)

####################Algeria site
lat_index <- which.min(abs(as.numeric(lat) - 28.7))
lon_index <- which.min(abs(as.numeric(lon) - 3.75))
Algeria_1420<- pcwd_annmax_1420[lon_index,lat_index,1:431]
Algeria_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
# Determine the maximum length between the two lists
length(Algeria_1420) <- max_length
# Prepend NAs to the shorter list
Algeria_1850 <- c(rep(NA, prepend_list2), Algeria_1850)
Algeria_df <- data.frame(time, Algeria_1420, Algeria_1850)

####################Antarctica site
lat_index <- which.min(abs(as.numeric(lat) - -81.5))
lon_index <- which.min(abs(as.numeric(lon) - 55))
Antarctica_1420<- pcwd_annmax_1420[lon_index,lat_index,1:431]
Antarctica_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
# Determine the maximum length between the two lists
length(Antarctica_1420) <- max_length
# Prepend NAs to the shorter list
Antarctica_1850 <- c(rep(NA, prepend_list2), Antarctica_1850)
Antarctica_df <- data.frame(time, Antarctica_1420, Antarctica_1850)

####################Czechia site
lat_index <- which.min(abs(as.numeric(lat) - 50.4))
lon_index <- which.min(abs(as.numeric(lon) - 13.75))
Czechia_1420<- pcwd_annmax_1420[lon_index,lat_index,1:431]
Czechia_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
# Determine the maximum length between the two lists
length(Czechia_1420) <- max_length
# Prepend NAs to the shorter list
Czechia_1850 <- c(rep(NA, prepend_list2), Czechia_1850)
Czechia_df <- data.frame(time, Czechia_1420, Czechia_1850)

####################Brazil site
lat_index <- which.min(abs(as.numeric(lat) - -7))
lon_index <- which.min(abs(as.numeric(lon) - -52.5))
Brazil_1420<- pcwd_annmax_1420[lon_index,lat_index,1:431]
Brazil_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
# Determine the maximum length between the two lists
length(Brazil_1420) <- max_length
# Prepend NAs to the shorter list
Brazil_1850 <- c(rep(NA, prepend_list2), Brazil_1850)
Brazil_df <- data.frame(time, Brazil_1420, Brazil_1850)


```

```{r plot multiple stations for both sets}

ggplot() +
  geom_line(data = Russia_df, aes(x = time, y = Russia_1420, color = "Russia 1420")) +
  geom_line(data = Russia_df, aes(x = time, y = Russia_1850, color = "Russia 1850")) +
  geom_line(data = Algeria_df, aes(x = time, y = Algeria_1420, color = "Algeria 1420")) +
  geom_line(data = Algeria_df, aes(x = time, y = Algeria_1850, color = "Algeria 1850")) +
    geom_line(data = Antarctica_df, aes(x = time, y = Antarctica_1420, color = "Antarctica 1420")) +
  geom_line(data = Antarctica_df, aes(x = time, y = Antarctica_1850, color = "Antarctica 1850")) +
    geom_line(data = Brazil_df, aes(x = time, y = Brazil_1420, color = "Brazil 1420")) +
  geom_line(data = Brazil_df, aes(x = time, y = Brazil_1850, color = "Brazil 1850")) +
    geom_line(data = Czechia_df, aes(x = time, y = Czechia_1420, color = "Czechia 1420")) +
  geom_line(data = Czechia_df, aes(x = time, y = Czechia_1850, color = "Czechia 1850")) +
  theme_classic() +
  ylim(0, max(Brazil_df$Brazil_1850)) +
  xlim(min(Russia_df$time), max(Russia_df$time)) +
  labs(
    x = "Year", 
    y = "Potential cumulative water deficit (mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Russia 1420" = "red", "Russia 1850" = "darkred","Algeria 1420" = "blue", "Algeria 1850" = "darkblue", "Antarctica 1420" = "darkslategray4", "Antarctica 1850" = "darkslategray", "Brazil 1420" = "green", "Brazil 1850" = "darkgreen", "Czechia 1420" = "yellow", "Czechia 1850" = "gold"))

ggsave("stations_allyears.tiff", width = 20, height = 14, units = "cm", dpi=320)


```

```{r plot multiple stations for both sets but around 1850}

ggplot() +
  geom_line(data = Russia_df, aes(x = time, y = Russia_1420, color = "Russia 1420")) +
  geom_line(data = Russia_df, aes(x = time, y = Russia_1850, color = "Russia 1850")) +
  geom_line(data = Algeria_df, aes(x = time, y = Algeria_1420, color = "Algeria 1420")) +
  geom_line(data = Algeria_df, aes(x = time, y = Algeria_1850, color = "Algeria 1850")) +
    geom_line(data = Antarctica_df, aes(x = time, y = Antarctica_1420, color = "Antarctica 1420")) +
  geom_line(data = Antarctica_df, aes(x = time, y = Antarctica_1850, color = "Antarctica 1850")) +
    geom_line(data = Brazil_df, aes(x = time, y = Brazil_1420, color = "Brazil 1420")) +
  geom_line(data = Brazil_df, aes(x = time, y = Brazil_1850, color = "Brazil 1850")) +
    geom_line(data = Czechia_df, aes(x = time, y = Czechia_1420, color = "Czechia 1420")) +
  geom_line(data = Czechia_df, aes(x = time, y = Czechia_1850, color = "Czechia 1850")) +
  theme_classic() +
  ylim(0, max(Brazil_df$Brazil_1850)) +
  xlim(1800, 1900) +
  labs(
    x = "Year", 
    y = "Potential cumulative water deficit (mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Russia 1420" = "red", "Russia 1850" = "darkred","Algeria 1420" = "blue", "Algeria 1850" = "darkblue", "Antarctica 1420" = "darkslategray4", "Antarctica 1850" = "darkslategray", "Brazil 1420" = "green", "Brazil 1850" = "darkgreen", "Czechia 1420" = "yellow", "Czechia 1850" = "gold"))

ggsave("stations_overlap.tiff", width = 20, height = 14, units = "cm", dpi=320)
```
